---
title: "Determinants of Insurance Premiums: Lifestyle Choices vs. Demographic Factors"
author: "Varenya Katta, Raphael Karamagi, Tim Jun, Oat Twinvitoo"
format: pdf
geometry: margin=1in
header-includes:
  - \usepackage{float}
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(tidyverse)
library(knitr)
library(kableExtra)
library(gridExtra)
library(broom)

# Load data
insurance_raw <- read.csv("data/medical_insurance.csv")

# 1. Define all relevant columns
relevant_cols <- c(
  "annual_premium", "age", "sex", "education", "income", 
  "marital_status", "household_size", "region", "urban_rural", 
  "bmi", "smoker", "alcohol_freq"
)

# 2. Select Columns & Filter Zeros
insurance <- insurance_raw |>
  select(all_of(relevant_cols)) |>
  filter(annual_premium > 0)

# 3. Factor Levels 
insurance <- insurance |>
  mutate(
    smoker = factor(smoker, levels = c("Never", "Former", "Current")),
    alcohol_freq = factor(alcohol_freq, levels = c("None", "Occasional", "Weekly", "Daily")),
    sex = factor(sex),
    education = factor(education),
    marital_status = factor(marital_status),
    region = factor(region),
    urban_rural = factor(urban_rural)
  )

insurance <- na.omit(insurance)
```

# 1. Introduction
In 2023, healthcare spending in the United States increased by 7.5%, totaling 4.9 trillion USD [8]. This parallel increase in healthcare spending and health insurance costs has exceeded the growth rates of wages and inflation, making it increasingly difficult for individuals to afford coverage [7].

However, health insurance costs are not uniform; they vary significantly from one individual to another based on risk profiles. Legal frameworks allow insurers to adjust premiums based on specific factors such as age, location, and tobacco use [6]. Yet, individuals with pre-existing health risks or specific lifestyle habits may self-select into more expensive plans. We can broadly categorize the factors driving healthcare costs into two distinct groups: Modifiable Lifestyle Choices (MLC) (behaviors an individual can change) and Baseline Demographic/Socioeconomic Factors (DRF) (fixed characteristics).

While many studies investigate these variables individually, this study attempts to fill a research gap by comparing the relative explanatory power of these two categories using a large-scale dataset. Specifically, does one category—lifestyle or demographics—have a more substantial effect on insurance cost variance? The results have implications for both policy and personal finance: if costs are driven by MLC, individuals can lower premiums by modifying behavior; if driven by DRF, systemic policy changes may be required.

Data Description We utilize the Medical Insurance Cost Prediction dataset from Kaggle [1], comprising approximately 100,000 observations. The constituent data was derived from CDC, HHS, and CMS logic [1]. The response variable is the Annual Premium (USD). We analyze two sets of predictors:

Modifiable Lifestyle Choices (MLC): Smoking status (Current, Former, Never), Alcohol consumption frequency, and Body Mass Index (BMI).

Baseline Demographics (DRF): Age, Sex, Education level, Income, Marital Status, and Household Size.



# 2. Exploratory Data Analysis

## 2.1 Summary Statistics

Table 1 summarizes the central tendencies of the continuous variables. The substantial difference between the median (463.58 USD) and mean (582.32 USD) premium indicates a right-skewed distribution, suggesting the presence of high-cost outliers.
```{r}
#| label: tbl-summary
insurance |>
  summarise(
    `Average Age` = mean(age),
    `Average BMI` = mean(bmi),
    `Median Premium` = median(annual_premium),
    `Mean Premium` = mean(annual_premium),
    `Max Premium` = max(annual_premium)
  ) |>
  kable(caption = "Summary of Numerical Variables", digits = 2) |>
  kable_styling(latex_options = "hold_position")
```

## 2.2 Distribution and Transformation

Linear regression assumes approximately normal residuals with constant variance. When the response variable is heavily right-skewed, these assumptions are typically violated. As shown in Figure 1 (Left), the raw annual premium distribution exhibits substantial right-skewness. We therefore applied a log-transformation, ln(Premium), which produces a more symmetric distribution (Figure 1, Right). This transformation typically improves residual normality and homoscedasticity (equal variances), which we verify in Section 4.1.

```{r}
# Fit Models
m1_base <- lm(log(annual_premium) ~ age + sex + education + income + marital_status + household_size + region + urban_rural, data = insurance)

m2_life <- lm(log(annual_premium) ~ smoker + alcohol_freq + bmi, data = insurance)

m3_full <- lm(log(annual_premium) ~ age + sex + education + income + marital_status + household_size + region + urban_rural + bmi + smoker + alcohol_freq + age*smoker, data = insurance)
```

```{r}
#| label: fig-distribution
#| fig-cap: "Distribution of annual premiums before and after log-transformation"
#| #| fig-pos: 'H'

p1 <- ggplot(insurance, aes(x = annual_premium)) +
  geom_histogram(bins = 50, fill = "orange", color = "white") +
  labs(title = "Raw Distribution (Skewed)", 
       x = "Annual Premium ($)", y = "Count") +
  theme_minimal()

p2 <- ggplot(insurance, aes(x = log(annual_premium))) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(title = "Log-Transformed (Normal)", 
       x = "Log(Premium)", y = "Count") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

## 2.3 Visualizing the "Smoking Penalty" and Interaction
We hypothesized that the cost burden of smoking might vary by age. Figure 2 plots Age against Log-Premium, stratified by smoking status.

The Green Line (Current Smokers) shows a higher intercept, indicating a higher baseline cost. More importantly, the slope appears steeper than that of the Red Line (Never Smokers). This suggests a potential interaction effect: the financial penalty for smoking compounds as the individual ages. We will test this formally in our regression models.

```{r}
#| label: fig-interaction
#| fig-pos: 'H'
#| fig-cap: "Log-premium by age, stratified by smoking status. The steeper slope for current smokers suggests an interaction effect."


ggplot(insurance, aes(x = age, y = log(annual_premium), color = smoker)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", se = FALSE, size = 1.5) +
  labs(title = "Interaction: Smoking Costs Increase Faster with Age",
       x = "Age of Individual", 
       y = "Log(Annual Premium)", 
       color = "Smoking Status") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")
```


# 3. Methodology

To isolate the variance explained by lifestyle choices versus demographics, we employ Hierarchical Multiple Linear Regression [3]. This approach allows us to enter variables in blocks to determine the incremental variance explained by each category. We utilize Analysis of Variance (ANOVA) using Partial F-tests to determine if the addition of a block of variables significantly reduces the Sum of Squared Errors (SSE) compared to a reduced model.

We fit three distinct models:

Model 1: Baseline Demographics Accounts only for fixed factors the individual cannot easily change. $$ \ln(Y) = \beta_0 + \beta_{Age} + \beta_{Sex} + \beta_{Region} + \beta_{Income} + \dots + \epsilon $$

Model 2: Lifestyle Only Accounts only for modifiable behaviors. $$ \ln(Y) = \beta_0 + \beta_{Smoker} + \beta_{Alcohol} + \beta_{BMI} + \epsilon $$

Model 3: Full Model (Demographics + Lifestyle) Combines both categories and includes the interaction term observed in EDA. $$ \ln(Y) = \text{Model 1} + \text{Model 2} + (\beta_{Age} \times \beta_{Smoker}) + \epsilon $$

Justification and Assumptions

We chose Ordinary Least Squares (OLS) regression on the log-transformed outcome. OLS provides highly interpretable coefficients, allowing us to quantify the precise percentage change in premiums associated with each lifestyle factor. The log-transformation addresses the right-skewness identified in EDA.

To ensure our results are valid, we will assess four key assumptions in the Results section:

1. Linearity: The relationship between predictors and the log-response is linear
2. Independence: We assume observations are independent (i.e., one individual's premium doesn't influence another's)
3. Equal Variance: The variance of residuals is constant across fitted values
4. Normality: The residuals are normally distributed

# 4. Results

## 4.1 Diagnostics and Assumptions Checking
To ensure the validity of our inference, we examined the residual plots for the full Model (Figure 3).

- Linearity & Equal Variance: The Residuals vs Fitted plot (Left) shows a random cloud of points around zero with no distinct pattern (e.g., no funnel shape). This indicates that the assumptions of linearity and constant variance are reasonably satisfied.

- Normality of Residuals: (Right) shows the Normal Q-Q plot. The points (circles) adhere closely to the diagonal line for a large number of observations. However, the tails deviate upward. This  behavior indicates that our model slightly underestimates the premiums for the most extreme, high-cost outliers. Given the large sample size ($N \approx 100,000$), the Central Limit Theorem ensures that our coefficient estimates remain unbiased despite this deviation.


```{r}
#| label: fig-diagnostics
#| #| fig-pos: 'H'
#| #| fig-cap: "Diagnostic plots for the Full Model. Left: Residuals vs. Fitted values. Right: Normal Q-Q plot."


par(mfrow = c(1, 2))
plot(m3_full, which = 1, main = "Residuals vs Fitted")
plot(m3_full, which = 2, main = "Normal Q-Q")


```

## 4.2 ANOVA Model Comparison

To test whether the addition of specific variable categories significantly improves the model, we utilized ANOVA F-tests.


```{r}
# Output ANOVA Table 1
anova(m1_base, m3_full) |>
  kable(caption = "ANOVA Comparison: Demographics vs. Full Model", digits = 2) |>
  kable_styling(latex_options = "hold_position")

# Output ANOVA Table 2
anova(m2_life, m3_full) |>
  kable(caption = "ANOVA Comparison: Lifestyle vs. Full Model", digits = 2) |>
  kable_styling(latex_options = "hold_position")
```

**Testing the Addition of Demographic Factors**
**H(null)**: \beta_{Age} = \beta_{Sex} =  \beta_{Region} = \beta_{Income} = \beta_{MaritalStatus} = \beta_{Urban/Rural} = \beta_{Education} = 0
**H(alt)**: At least one of  \beta_{Age}, \beta_{Sex}, \beta_{Region}, \beta_{Income}, \beta_{MaritalStatus}, \beta_{Urban/Rural},  \beta_{Education} ≠ 0
We compared the Lifestyle-only model against the Full model to see if adding Demographics significantly reduced the squared residuals. The F-test yielded an F-statistic of 461.19 and a p-value displaying 0, clearly less than <0.05. We reject the null hypothesis that all demographic coefficients are zero. Therefore, variation at least one of the socioeconomic/demographic factors seems to help explain the variation in the log(monthly premium); the addition of DRF helps reduce the squared residuals and provide statistically significant explanatory power.

**Testing the Addition of Modifiable Lifestyle Factors** 
**H(null)**: \beta_{Smoker} = \beta_{Alcohol} =  \beta_{bmi} = 0
**H(alt)**: At least one of \beta_{Smoker},  \beta_{Alcohol},  \beta_{bmi} ;  ≠ 0
We compared the Demographic-only model against the Full model to see if adding Lifestyle factors improved the fit. The F-test yielded an F-statistic of 128.88 and a p-value displaying 0, clearly less than <0.05. We reject the null hypothesis that lifestyle coefficients are zero. Therefore, variation at least one of the modifiable life choices seems to help explain the variation in the log(annual premium); This confirms that modifiable choices (Smoking, BMI) also significantly explain variation in premiums.

The ANOVA tests conclude that the addition of MLC parameters to a DRF model may help explain the variation in the log(annual premium). However, as the degrees of freedom are not the same for each ANOVA test, the F value cannot be used directly to compare the impact of the added Demographic/Modifiable variables.

## 4.3 Comparison of R² Values

While ANOVA tests for statistical significance, comparing R² values allows us to assess the practical explanatory power (effect size).

```{r}
R21 <- summary(m1_base)$r.squared
R22 <- summary(m2_life)$r.squared
R23 <- summary(m3_full)$r.squared

dfr2 <- data.frame(
  Category = c("R\u00B2 Demographic", "R\u00B2 Lifestyle", " R\u00B2 Full Model"),
  R2_Value = c(R21, R22, R23)
)

kable(dfr2, caption = "R\u00B2 Values of Demographic, Lifestyle, and Full Model")

R231 <- R23 - R21
R232 <- R23 - R22

dfr2c <- data.frame(
  Category = c("ΔR\u00B2 (Full Model - Demographic)", "ΔR\u00B2 (Full Model - Lifestyle)"),
  R2_Value = c(R231, R232)
)

kable(dfr2c, caption = "ΔR\u00B2 of Demographic and Lifestyle Model Compared to Full Model")

```

The R² values are generally low across all models (0.0249 to 0.0596), indicating that the variables available in this dataset explain only a small fraction of the total variance in premiums.

However, we can compare the relative impact of the two categories. Adding Lifestyle variables to the demographic model increased the R² by roughly 0.0347. Conversely, adding Demographic variables to the lifestyle model increased the R² by only 0.0255. This suggests that modifiable lifestyle factors have a stronger marginal effect on predicting insurance costs than fixed demographic factors.


## 4.4 Regression Coefficients
Table 5 presents the coefficients for the Full Model. Since the response is log-transformed, coefficients ($\beta$) can be interpreted as an approximate $100 \times \beta \%$ change in the premium for small values.

```{r}
# Tidy results for the main lifestyle factors
tidy(m3_full) |>
  filter(str_detect(term, "bmi|smoker|alcohol|age:smoker")) |>
  mutate(
    term = str_replace_all(term, "smokerCurrent", "Smoker: Current"),
    term = str_replace_all(term, "smokerFormer", "Smoker: Former"),
    term = str_replace_all(term, "alcohol_freqWeekly", "Alcohol: Weekly"),
    term = str_replace_all(term, "alcohol_freqOccasional", "Alcohol: Occasional"),
    term = str_replace_all(term, "alcohol_freqDaily", "Alcohol: Daily"),
    term = str_replace_all(term, "age:smokerCurrent", "Interaction: Age * Smoker"),
    term = str_replace_all(term, "bmi", "BMI"),
    estimate = round(estimate, 5),
    p.value = scales::pvalue(p.value)
  ) |>
  select(Predictor = term, Estimate = estimate, `Std. Error` = std.error, `P-Value` = p.value) |>
  kable(caption = "Coefficients for Modifiable Lifestyle Factors (Model 2)") |>
  kable_styling(latex_options = "hold_position")
```

Key Findings: 

1. Smoking: Being a "Current Smoker" has a large positive coefficient ($0.221$). Holding other factors constant, current smokers pay approximately 22% more in premiums than never-smokers at the baseline age.

2. BMI: The coefficient for BMI is positive and statistically significant ($p < 0.001$). Higher body mass index is reliably associated with higher insurance costs.

3. Alcohol: The p-values for alcohol frequency are $> 0.05$. This suggests that, unlike smoking and BMI, self-reported alcohol frequency is not a statistically significant predictor of insurance premiums in this dataset.

4. Interaction (Age $\times$ Smoker): The significant interaction term ($p = 0.002$) confirms our EDA finding. This positive coefficient ($\beta = 0.00087$)) means the premium gap between smokers and non-smokers widens as age increases. For every additional year of age, the "smoker penalty" increases.


# 5. Discussion




# 6. References

[1] Kaggle. (2024). Medical Insurance Cost Prediction Dataset.
[2] Wooldridge, J. M. (2015). Introductory Econometrics.
[3] UVa Library. (2023). Hierarchical Linear Regression. https://library.virginia.edu/data/articles/hierarchical-linear-regression
[4] American Medical Association. (2024). National Health Expenditure Data. https://www.ama-assn.org/
[5] Bureau of Labor Statistics. (2023). Consumer Expenditure Survey: Health Insurance.
[6] HealthCare.gov. (2024). How Insurance Companies Set Health Premiums. https://www.healthcare.gov/how-plans-set-your-premiums/
[7] Kaiser Family Foundation. (2023). Employer Health Benefits Survey. https://www.kff.org/
[8] Centers for Medicare & Medicaid Services, Office of the Actuary. “National Health Expenditure (NHE) Fact Sheet: Historical NHE, 2023.” U.S. Department of Health & Human Services