---
title: "Determinants of Insurance Premiums: Lifestyle Choices vs. Demographic Factors"
author: "Varenya Katta, Raphael Karamagi, Tim Jun, Oat Twinvitoo"
format: pdf
geometry: margin=1in
header-includes:
  - \usepackage{float}
execute:
  echo: false
  warning: false
  message: false
  fig.width: 4
  fig.height: 3
---

```{r setup}
#| include: false
library(tidyverse)
library(knitr)
library(kableExtra)
library(gridExtra)
library(broom)

# Load data
insurance_raw <- read.csv("data/medical_insurance.csv")

# 1. Define all relevant columns
relevant_cols <- c(
  "annual_premium", "age", "sex", "education", "income", 
  "marital_status", "household_size", "region", "urban_rural", 
  "bmi", "smoker", "alcohol_freq"
)

# 2. Select Columns & Filter Zeros
insurance <- insurance_raw |>
  select(all_of(relevant_cols)) |>
  filter(annual_premium > 0)

# 3. Factor Levels 
insurance <- insurance |>
  mutate(
    smoker = factor(smoker, levels = c("Never", "Former", "Current")),
    alcohol_freq = factor(alcohol_freq, levels = c("None", "Occasional", "Weekly", "Daily")),
    sex = factor(sex),
    education = factor(education, levels = c("No HS", "HS", "Some College", "Bachelors", "Masters", "Doctorate")),
    marital_status = factor(marital_status),
    region = factor(region),
    urban_rural = factor(urban_rural)
  )

insurance <- na.omit(insurance)

glimpse(insurance)
```

# 1. Introduction

# 1.1 Context and Research Question

In 2023, healthcare spending in the United States increased by 7.5%, totaling 4.9 trillion USD [8]. This increase in healthcare spending paralells increases in health insurance costs. As of recent health insurance costs have exceeded the growth rates of wages and inflation, making it increasingly difficult for individuals to afford coverage [7].

However, health insurance vary from individual to individual; according to the Affordable Care Act, factors such as age, location and tobacco use are legally allowed to increase the cost of one's premium [6]. Further, individuals with other pre-existing health risks may also select costlier healthcare plans. Overall, we can categorize factors that potentially drive healthcare cost into two categories: Modifiable Lifestyle Choices (MLC) (behaviors an individual can change) and Baseline Demographic/Socioeconomic Factors (DRF) (fixed characteristics).

This study will attempt to compare the extent of variability attributable to each of these two categories: Does one category have a more substantial effect on insurance cost variance? The results may indicate how institutions and individuals can make healthcare more affordable. If insurance cost variance is heavily driven by MLC, customers can lower prices by modifying behaviors. However, if insurance cost is heavily driven by DRF, policy changes may be required to lower costs.

Research Question: **To what extent can variation in insurance premiums be attributed to Baseline Demographic Factors (DRF) versus Modifiable Lifestyle Factors (MLC)**

**Dataset:** We use the Medical Insurance Cost Prediction dataset from Kaggle [1]. The dataset contains 100,000 observations including demographics (Age, Sex, Income) and lifestyle indicators (Smoking status, BMI, Alcohol frequency). Data was derived from three main sources: Centers for Disease Control & Prevention (CDC), Department of Health & Human Services (HHS), and Centers for Medicare & Medicaid Services (CMS). The response variable is the annual healthcare premium cost.

Modifiable Lifestyle Choices (MLC): Smoking status (Current, Former, Never), Alcohol consumption frequency, and Body Mass Index (BMI).

Baseline Demographics (DRF): Age, Sex, Education level, Income, Marital Status, and Household Size.

# 1.2 Exploratory Data Analysis

Table 1 provides summary statistics of the response variable (Annual Premium), namely, median, mean, minimum and maximum. As the mean premium, at 582.32 USD, is substantially higher than the median, at 463.58 USD, we can infer a right-skewed distribution. Further, the high value of the maximum premium, at 10962.55 USD, especially in comparison to the minimum premium value of 211.67 USD, may imply the presence of high outliers. 

```{r}
insurance |>
  summarise(
    `Median Premium` = median(annual_premium),
    `Mean Premium` = mean(annual_premium),
    `Min Premium` = min(annual_premium),
    `Max Premium` = max(annual_premium)
  ) |>
  kable(caption = "Summary Statistics of Annual Premium Costs", digits = 2) |>
  kable_styling(latex_options = "hold_position")
```

# 2. Methodology

```{r}
# Fit Models
p3_full <- lm((annual_premium) ~ age + sex + education + income + marital_status + household_size + region + urban_rural + bmi + smoker + alcohol_freq, data = insurance)
```

```{r}
# Fit Models
m1_base <- lm(log(annual_premium) ~ age + sex + education + income + marital_status + household_size + region + urban_rural, data = insurance)

m2_life <- lm(log(annual_premium) ~ smoker + alcohol_freq + bmi, data = insurance)

m3_full <- lm(log(annual_premium) ~ age + sex + education + income + marital_status + household_size + region + urban_rural + bmi + smoker + alcohol_freq, data = insurance)
```

To isolate the variance explained by lifestyle choices versus demographics, we employ Hierarchical Multiple Linear Regression [3]. This approach allows us to enter variables in blocks to determine the unique variance explained by each category. We utilize Analysis of Variance (ANOVA) using Partial F-tests to determine if the addition of a block of variables significantly reduces the Sum of Squared Errors (SSE) compared to a reduced model.

We fit three distinct models:

Model 1: Baseline Demographics Accounts only for fixed factors the individual cannot easily change. $$ \ln(Y) = \beta_0 + \beta_{Age} x_{Age} + \beta_{Sex} x_{Sex} + \beta_{Region} x_{Region} + \beta_{Income} x_{Income} + \cdots + \epsilon $$

Model 2: Lifestyle Only Accounts only for modifiable behaviors. $$ \ln(Y) = \ln(Y) = \beta_0 + \beta_{Smoker} x_{Smoker} + \beta_{Alcohol} x_{Alcohol} + \beta_{BMI} x_{BMI} + \epsilon
 $$

Model 3: Full Model (Demographics + Lifestyle) Combines both categories and includes the interaction term observed in EDA. $$ \ln(Y) = \text{Model 1} + \text{Model 2} + \epsilon $$

While ANOVA tests for statistical significance, it does not quantify effect size. Therefore, we also compare the change in Adjusted R² between the partial models (DRF vs. Full and MLC vs. Full). This allows us to assess which set of variables has a more substantial practical effect on reducing the squared residuals.

# 2.1 Log Transformation
The use of a log transformation addresses the requirement of normality in the residuals required by Least Squares Regression [2]. When the outcome is not transformed, the distribution is right skewed (Figure 1, Left). Applying a log transformation [2] to the annual premium produced a more symmetric distribution (Figure 1, Right), allowing the normality requirement to be satisfied.

```{r}
#| fig.cap: "Left: Right-skewed distribution of raw premiums. Right: Approximate normal distribution after log-transformation."
#| fig.height: 3
#| fig.width: 8

p1 <- ggplot(insurance, aes(x = annual_premium)) + 
  geom_histogram(bins = 50, fill = "orange", color = "white") +
  labs(title = "Raw Distribution", x = "Annual Premium ($)", y = "Count") +
  theme_minimal()

p2 <- ggplot(insurance, aes(x = log(annual_premium))) + 
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(title = "Log-Transformed", x = "Log(Annual Premium)", y = "Count") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```


# 2.2 Justification and Other Assumptions

We chose Ordinary Least Squares (OLS) regression on the log-transformed outcome. OLS provides highly interpretable coefficients, allowing for the quantification of percentage changes in premiums associated with each lifestyle and demographic variable.

To ensure our inference is valid, we must verify the assumptions for both OLS regression and the ANOVA F-tests used for model comparison. These assumptions largely overlap and will be assessed in the Results section using residual diagnostics:

1. Linearity: The relationship between predictors and the log-response must be linear. We will assess this by examining the Residuals vs. Fitted plot for random scatter.

2. Independence: Observations must be independent. While we assume this holds for the analysis, we note that the voluntary nature of the survey data may introduce selection bias.

3. Equal Variance (Homoscedasticity): The variance of residuals must be constant across all fitted values. We will verify this by checking for a "funnel shape" in the Residuals vs. Fitted plot.

4. Normality (OLS & ANOVA): Both OLS and ANOVA require that the error terms are normally distributed. We will assess this using a Normal Q-Q plot.

5. Homogeneity of Variance (ANOVA): Similar to OLS, ANOVA assumes variance is consistent across groups. This is satisfied if the Equal Variance assumption for the full model holds.


# 3. Results

## 3.1 Diagnostics and Assumptions Checking
To ensure the validity of our inference, we examined the residual plots for the full Model (Figure 3) as well as data collection procedures.

- Linearity & Equal Variance: The Residuals vs Fitted plot (Left) shows a random cloud of points around zero with no distinct pattern (e.g., no funnel shape). This indicates that the assumptions of linearity and constant variance are reasonably satisfied.

- Independence: One key source of this dataset [1] is the National Health Interview survey; this survey is conducted on a voluntary basis. Consequently, there is a self-selection bias that may affect the ability to fully assume independence for this study. Further, confounding variables may exist. For instance, variables such as location may be affecting bmi and smoking.

- Normality of Residuals: As shown earlier in the residual histograms, the normality of the residuals for each of the three models is roughly fulfilled. Further, figure 4 (Right) shows the Normal Q-Q plot for the full model. The points (circles) adhere closely to the diagonal line for the majority of observations. However, the tails deviate upward. This "heavy-tailed" behavior indicates that our model slightly underestimates the premiums for the most extreme, high-cost outliers. Given the large sample size ($N \approx 100,000$), the Central Limit Theorem ensures that our coefficient estimates remain unbiased despite this deviation [2].


```{r}
#| label: fig-diagnostics
#| fig.cap: "Diagnostic plots for the Full Model. Left: Residuals vs. Fitted values showing constant variance. Right: Normal Q-Q plot showing approximate normality."
#| fig.height: 3.5

par(mfrow = c(1, 2))
plot(m3_full, which = 1, main = "Residuals vs Fitted")
plot(m3_full, which = 2, main = "Normal Q-Q")
```

## 3.2 ANOVA Model Comparison
To test whether the addition of specific variable categories significantly improves the model, we utilized ANOVA F-tests.


```{r}
# Output ANOVA Table 1
anova(m1_base, m3_full) |>
  kable(caption = "ANOVA Comparison: Demographics vs. Full Model", digits = 2) |>
  kable_styling(latex_options = "hold_position")

# Output ANOVA Table 2
anova(m2_life, m3_full) |>
  kable(caption = "ANOVA Comparison: Lifestyle vs. Full Model", digits = 2) |>
  kable_styling(latex_options = "hold_position")
```


**Testing the Addition of Demographic Factors**

**H(null):** $\beta_{Age} = \beta_{Sex} =  \beta_{Region} = \beta_{Income} = \beta_{MaritalStatus} = \beta_{Urban/Rural} = \beta_{Education}  = 0$ (Demographics add no explanatory power)

**H(alt):** At least one of  $\beta_{Age}, \beta_{Sex}, \beta_{Region}, \beta_{Income}, \beta_{MaritalStatus}, \beta_{Urban/Rural},  \beta_{Education} ≠ 0$ 

We compared the Lifestyle-only model against the Full model to see if adding Demographics significantly reduced the squared residuals. The F-test yielded an F-statistic of 613.24 and a p-value <0.001. We reject the null hypothesis and conclude that at least one of the coefficients of DRF has a statistically significant difference from zero. This means adding the DRF variables results in a higher R². This confirms that demographic factors (Age, Income, etc.) do provide statistically significant explanatory power. 

**Testing the Addition of Modifiable Lifestyle Factors** 

**H(null)**: $\beta_{Smoker} = \beta_{Alcohol} =  \beta_{BMI} = 0$ (Lifestyle Choices add no explanatory power)

**H(alt)**: At least one of $\beta_{Smoker},  \beta_{Alcohol},  \beta_{BMI}  ≠ 0$

We compared the Demographic-only model against the Full model to see if adding Lifestyle factors improved the fit. The F-test yielded an F-statistic of 141.92 and a p-value <0.001. We reject the null hypothesis and conclude that at least one of the coefficients of MLC has a statistically significant difference from zero. This means adding the MLC variables results in a higher R².  This confirms that modifiable choices (Smoking, BMI) also significantly explain variation in premiums.

The ANOVA tests conclude that the addition of MLC parameters to a DRF model helps explain the variation in the log(annual premium). However, as the degrees of freedom are not the same for each ANOVA test, the F value cannot be used directly to compare the practical impact of the added variables.

## 3.3 Comparison of R² Values

While ANOVA tests for statistical significance, comparing R² values allows us to assess the practical explanatory power (effect size).

```{r}
R21 <- summary(m1_base)$r.squared
R22 <- summary(m2_life)$r.squared
R23 <- summary(m3_full)$r.squared

dfr2 <- data.frame(
  Category = c("R² Demographic", "R² Lifestyle", "R² Full Model"),
  R2_Value = c(R21, R22, R23)
)

kable(dfr2, caption = "R² Values of Demographic, Lifestyle, and Full Model", digits = 4) |>
  kable_styling(latex_options = "hold_position")

R231 <- R23 - R21
R232 <- R23 - R22

dfr2c <- data.frame(
  Category = c("Change in R² (Full - Demographic)", "Change in R² (Full - Lifestyle)"),
  R2_Value = c(R231, R232)
)

kable(dfr2c, caption = "Marginal Contribution to R²", digits = 4) |>
  kable_styling(latex_options = "hold_position")
```

The R² values are generally low across all models (0.0249 to 0.0595), indicating that the variables available in this dataset explain only a small fraction of the total variance in premiums.

However, we can compare the relative impact of the two categories. The removal of MLC from the full model decreased the R² by roughly 0.0346. Conversely, the removal of DRF from the full model increased the R² by only 0.0254. This suggests that modifiable lifestyle factors have a stronger marginal effect on predicting insurance costs than fixed demographic factors.


## 3.4 Regression Coefficients
Table 5 presents the coefficients for bmi, smoker status, alcohol usage, age, and education level for the Full Model. These coefficients were specifically included as the p values associated were surprising or significant.

```{r}
# Tidy results for the Some Key Factors
tidy(m3_full) |>
  filter(str_detect(term, "bmi|smoker|alcohol|age|education|region")) |>
  mutate(
    term = str_replace_all(term, "smokerCurrent", "Smoker: Current"),
    term = str_replace_all(term, "smokerFormer", "Smoker: Former"),
    term = str_replace_all(term, "alcohol_freqWeekly", "Alcohol: Weekly"),
    term = str_replace_all(term, "alcohol_freqOccasional", "Alcohol: Occasional"),
    term = str_replace_all(term, "alcohol_freqDaily", "Alcohol: Daily"),
    term = str_replace_all(term, "bmi", "BMI"),
    term = str_replace_all(term, "age", "AGE"),
    term = str_replace_all(term, "educationHS", "Education: HS"),
    term = str_replace_all(term, "educationSome College", "Education: Some College"),
    term = str_replace_all(term, "educationBachelors", "Education: Bachelors"),
    term = str_replace_all(term, "educationMasters", "Education: Masters"),
    term = str_replace_all(term, "educationDoctorate", "Education: Doctorate"),
    term = str_replace_all(term, "regionEast", "Region: East"),
    term = str_replace_all(term, "regionNorth", "Region: North"),
    term = str_replace_all(term, "regionSouth", "Region: South"),
    term = str_replace_all(term, "regionWest", "Region: West"),
    estimate = round(estimate, 5),
    p.value = scales::pvalue(p.value)
  ) |>
  select(Predictor = term, Estimate = estimate, `Std. Error` = std.error, `P-Value` = p.value) |>
  kable(caption = "Coefficients for MLC and DRF variables (Model 3)") |>
  kable_styling(latex_options = "hold_position", font_size = 9)
```

Key Findings: 

1. Smoking: Being a current smoker and a former smoker has a p value that is statistically significant ($p < 0.01$). Notably a "Current Smoker" has a large positive coefficient ($0.263$). Holding other factors constant, current smokers have a log-premium that is 0.263 units higher than never-smokers.

2. BMI: The coefficient for BMI is positive and statistically significant ($p < 0.01$). The model predicts that for each increase in one BMI point, the log(annual premium) increases by 0.00562.

3. Alcohol: The p-values for alcohol frequency are $> 0.01$. This suggests that, unlike smoking and BMI, self-reported alcohol frequency is not a statistically significant predictor of log(annual premiums) in this dataset.

4. Age: p values were  ($p < 0.01$), indicating that change in age leads to a statistically significant change in log (annual  premiums). The model predicts that for each increase in one year of age point, the log(annual premium) increases by 0.0048.

5. Education: p values were  ($p > 0.01$) for all education levels. This suggests that education is not a statistically significant predictor of log(annual premiums) in this dataset. However, notably, there are significantly lower p values associated with holding a doctorate, at 0.026, compared to other education levels which are all $> 0.05$.

6. Region: p values were ($p > 0.01$) for all regions. This suggests that being in the north, east, south, and west does not lead to a significantly significant change in log(annual premium) compared to those who lived in central US.


# 4. Discussion

## 4.1 Results Interpretation

Our analysis indicates that both modifiable lifestyle choices (MLC) and demographic factors (DRF) significantly affect insurance premiums, though lifestyle choices appear to have a stronger practical impact. The "Lifestyle Only" model explained more variance than the "Demographics Only" model, and adding MLC variables to the model provided a larger boost to explanatory power than adding demographics.

The regression coefficients confirm that specific lifestyle choices carry a heavy financial weight. Current smokers face a log-premium penalty of ~0.26 compared to non-smokers, aligning with ACA provisions allowing tobacco surcharges. BMI also consistently increases costs. Conversely, alcohol consumption and most education levels showed no significant effect, which aligns with ACA regulations prohibiting discrimination based on health history. Surprisingly, our broad regional categories did not show significant deviations from the Central US, suggesting price variations might be more localized.

## 4.2 Methodological Limitations & Recommendations
Several issues limit the strength of these conclusions. 
First, some factors that are known to have been able to affect healthcare costs, such as plan types, are not included in the data set/analysis - like plan type (Gold/Silver) and deductibles, and these likely explain the low total R² values (<6%). 
Second, some of the data sources such as the National Health Interview Survey collects self reported data. Relying upon such data for factors such as BMI, smoking and alcohol consumption may be prone to error as individuals may want to underreport such stigmatized behaviors. 
Third, the model underestimates premiums for the individuals paying the most for insurance, as shown in the Q-Q plot's heavy tails.
Fourth, as the variables selected do not represent the totality of Modifiable Lifestyle Choices and Demographics, a general conclusion about these two factors may be inappropriate. 

To improve the study, more variables representing MLC and DRF should be included. It may also be helpful to test the interaction effects between MLC and DRF variables. Further, more categories for alcohol consumption may be helpful: Currently, there is no distinction made between an individual who has quit alcohol recently and another who has quit alcohol for a long duration of time. It would seem logical that a distinction in insurance premium between these two groups may exist. Finally, a dataset that does not rely on self-reported data should be found. We also recommend greater exploration of the education status, as while all p values were greater than 0.01; the p value associated with having a doctorate was still quite low at 0.026. Studies may also want to subdivide regions of the United States into more segments than utilized in this study (north, east, south, west, central), and investigate the more subdivided regions will have statistically significant health insurance costs.

## 4.3 Implications
Our results reveal that lifestyle choices and demographics contribute to the cost of insurance. The addition of lifestyle choice variables seem to help explain variability in log(annual insurance cost) more significantly when demographic factors were added. Despite this, the examined variables seem to explain very little variation in log(annual insurance costs), as indicated by the low R² value. Simultaneously, this implies that other variables not investigated in this study may better explain variations in health insurance. Consequently, further study is needed into the explanatory effect of modifiable lifestyle choice variables vs. demographic variables on insurance cost.


# 5. References

[1] Kaggle. (2024). Medical Insurance Cost Prediction Dataset.
[2] Wooldridge, J. M. (2015). Introductory Econometrics.
[3] UVa Library. (2023). Hierarchical Linear Regression. https://library.virginia.edu/data/articles/hierarchical-linear-regression
[4] American Medical Association. (2024). National Health Expenditure Data. https://www.ama-assn.org/
[5] Bureau of Labor Statistics. (2023). Consumer Expenditure Survey: Health Insurance.
[6] HealthCare.gov. (2024). How Insurance Companies Set Health Premiums. https://www.healthcare.gov/how-plans-set-your-premiums/
[7] Kaiser Family Foundation. (2023). Employer Health Benefits Survey. https://www.kff.org/
[8] Centers for Medicare & Medicaid Services, Office of the Actuary. “National Health Expenditure (NHE) Fact Sheet: Historical NHE, 2023.” U.S. Department of Health & Human Services