---
title: "Determinants of Insurance Premiums: Lifestyle Choices vs. Demographic Factors"
author: "Varenya Katta, Raphael Karamagi, Tim Jun, Oat Twinvitoo"
format: pdf
geometry: margin=1in
header-includes:
  - \usepackage{float}
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(tidyverse)
library(knitr)
library(kableExtra)
library(gridExtra)
library(broom)

# Load data
insurance_raw <- read.csv("data/medical_insurance.csv")

# 1. Define all relevant columns
relevant_cols <- c(
  "annual_premium", "age", "sex", "education", "income", 
  "marital_status", "household_size", "region", "urban_rural", 
  "bmi", "smoker", "alcohol_freq"
)

# 2. Select Columns & Filter Zeros
insurance <- insurance_raw |>
  select(all_of(relevant_cols)) |>
  filter(annual_premium > 0)

# 3. Factor Levels 
insurance <- insurance |>
  mutate(
    smoker = factor(smoker, levels = c("Never", "Former", "Current")),
    alcohol_freq = factor(alcohol_freq, levels = c("None", "Occasional", "Weekly", "Daily")),
    sex = factor(sex),
    education = factor(education),
    marital_status = factor(marital_status),
    region = factor(region),
    urban_rural = factor(urban_rural)
  )

insurance <- na.omit(insurance)
```

# 1. Introduction
Within 2023, healthcare spending in the United States increased by 7.5%, totaling to 4.9 trillion USD; this healthcare spending increase parallels increases in the cost of health insurance. Recent rates of change in health insurance premium cost has exceeded that of wages and inflation, making it harder for individuals to afford health insurance. 
However, health insurance costs vary from one individual to another; factors such as age, location and tobacco use are legally allowed to increase the cost of one's premium. 

However, individuals with other pre-existing health risks may themselves select costlier healthcare plans that provide greater coverage. Further, other variables may affect tobacco use. Overall, we can categorize factors that potentially drive healthcare cost into two categories: modifiable lifestyle choices and baseline demographic/socioeconomic factors. 

Many studies have investigated the role of these variables individually. This study attempts to fill a research gap by investigating how much premium prices can be attributed by baseline demographic/socioeconomic factors (DRF) versus modifiable lifestyle factors (MLC); does one category have a more substantial effect on insurance cost variance? The results indicate how institutions and individuals can make healthcare more affordable. For instance, if insurance cost variance is heavily driven by MLC, customers can lower prices by modifying behaviors. However, if insurance cost is heavily driven by DRF, policy changes may be required to lower costs.

Tthree variables representing MLC are considered: smoking status (current, former, never), alcohol consumption frequency (never, occasionally, weekly, daily), and BMI. We consider six variables representing DRF, age, sex, highest education level, income, marital status, and household size.

Dataset: We utilize the Medical Insurance Cost Prediction dataset from Kaggle [1]. The dataset contains approximately 100,000 observations including demographic details (Age, Sex, Income) and lifestyle indicators (Smoking status, BMI, Alcohol frequency). Constituent data was derived from three main sources: Centers for Disease Control and Prevention (CDC), U.S. Department of Health & Human Services (HHS), and Centers for Medicare & Medicaid Services (CMS). 

Research Question: **How much can premium prices be attributed by baseline demographic/socioeconomic factors (DRF) versus modifiable lifestyle factors (MLC)?**




# 2. Exploratory Data Analysis

## 2.1 Summary Statistics

Table 1 summarizes the central tendencies of the continuous variables. The substantial difference between the median (463.58 USD) and mean (582.32 USD) premium indicates a right-skewed distribution, suggesting the presence of high-cost outliers.
```{r}
#| label: tbl-summary
insurance |>
  summarise(
    `Average Age` = mean(age),
    `Average BMI` = mean(bmi),
    `Median Premium` = median(annual_premium),
    `Mean Premium` = mean(annual_premium),
    `Max Premium` = max(annual_premium)
  ) |>
  kable(caption = "Summary of Numerical Variables", digits = 2) |>
  kable_styling(latex_options = "hold_position")
```

## 2.2 Distribution and Transformation
Linear regression assumes approximately normal residuals with constant variance. When the response variable is heavily right-skewed, these assumptions are typically violated. As shown in Figure 1 (Left), the raw annual premium distribution exhibits substantial right-skewness. We therefore applied a log-transformation, ln(Premium), which produces a more symmetric distribution (Figure 1, Right). This transformation typically improves residual normality and homoscedasticity (equal variances), which we verify in Section 4.1.

```{r}
# Fit Models
m1_base <- lm(log(annual_premium) ~ age + sex + education + income + marital_status + household_size + region + urban_rural, data = insurance)

m2_life <- lm(log(annual_premium) ~ smoker + alcohol_freq + bmi, data = insurance)

m3_full <- lm(log(annual_premium) ~ age + sex + education + income + marital_status + household_size + region + urban_rural + bmi + smoker + alcohol_freq + age*smoker, data = insurance)
```

```{r}
#| label: fig-distribution
#| fig-cap: "Distribution of annual premiums before and after log-transformation"
#| #| fig-pos: 'H'

p1 <- ggplot(insurance, aes(x = annual_premium)) +
  geom_histogram(bins = 50, fill = "orange", color = "white") +
  labs(title = "Raw Distribution (Skewed)", 
       x = "Annual Premium ($)", y = "Count") +
  theme_minimal()

p2 <- ggplot(insurance, aes(x = log(annual_premium))) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(title = "Log-Transformed (Normal)", 
       x = "Log(Premium)", y = "Count") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

## 2.3 Visualizing the "Smoking Penalty" and Interaction
We hypothesized that the cost burden of smoking might vary by age. Figure 2 plots Age against Log-Premium, stratified by smoking status.

The Green Line (Current Smokers) shows a higher intercept, indicating a higher baseline cost. More importantly, the slope appears steeper than that of the Red Line (Never Smokers). This suggests a potential interaction effect: the financial penalty for smoking compounds as the individual ages. We will test this formally in our regression models.

```{r}
#| label: fig-interaction
#| fig-pos: 'H'
#| fig-cap: "Log-premium by age, stratified by smoking status. The steeper slope for current smokers suggests an interaction effect."


ggplot(insurance, aes(x = age, y = log(annual_premium), color = smoker)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", se = FALSE, size = 1.5) +
  labs(title = "Interaction: Smoking Costs Increase Faster with Age",
       x = "Age of Individual", 
       y = "Log(Annual Premium)", 
       color = "Smoking Status") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")
```


# 3. Methodology

To isolate the variance explained by lifestyle choices versus demographics, we employ Hierarchical Multiple Linear Regression [3]. This approach allows us to enter variables in blocks to determine the incremental variance explained by each category.

We fit three distinct models:

Model 1: Baseline Demographics Accounts only for fixed factors the individual cannot easily change. $$ \ln(Y) = \beta_0 + \beta_{Age} + \beta_{Sex} + \beta_{Region} + \beta_{Income} + \dots + \epsilon $$

Model 2: Lifestyle Only Accounts only for modifiable behaviors. $$ \ln(Y) = \beta_0 + \beta_{Smoker} + \beta_{Alcohol} + \beta_{BMI} + \epsilon $$

Model 3: Full Model (Demographics + Lifestyle) Combines both categories and includes the interaction term observed in EDA. $$ \ln(Y) = \text{Model 1} + \text{Model 2} + (\beta_{Age} \times \beta_{Smoker}) + \epsilon $$

Justification and Diagnostics

We chose Ordinary Least Squares (OLS) regression on the log-transformed outcome. OLS provides highly interpretable coefficients, allowing us to quantify the precise percentage change in premiums associated with each lifestyle factor. The log-transformation addresses the right-skewness identified in EDA.

To ensure our results are valid, we will assess three key assumptions in the Results section:

Linearity & Homoscedasticity: Using Residuals vs. Fitted plots to ensure errors have constant variance.

Normality: Using Q-Q plots to ensure residuals follow a normal distribution.

Independence: 

# 4. Results

## 4.1 Diagnostics and Assumptions Checking
To ensure the validity of our inference, we examined the residual plots for the full Model.

- Linearity & Equal Variance: The Residuals vs Fitted plot (Figure 3, Left) shows a random cloud of points around zero, indicating these assumptions are met.

- Independence: One key source of this dataset is the National Health Interview survey; this survey is conducted on a voluntary basis. Consequently, there is a self-selection bias that may affect the ability to fully assume independence for this study.

- Normality of Residuals: Figure 4 (Right) shows the Normal Q-Q plot. The points (circles) adhere closely to the diagonal line for the majority of observations. However, the tails deviate upward. This "heavy-tailed" behavior indicates that our model slightly underestimates the premiums for the most extreme, high-cost outliers. Given the large sample size ($N \approx 100,000$), the Central Limit Theorem ensures that our coefficient estimates remain unbiased despite this deviation [2].


```{r}
#| label: fig-diagnostics
#| #| fig-pos: 'H'
#| #| fig-cap: "Diagnostic plots for the Full Model. Left: Residuals vs. Fitted values. Right: Normal Q-Q plot."


par(mfrow = c(1, 2))
plot(m3_full, which = 1, main = "Residuals vs Fitted")
plot(m3_full, which = 2, main = "Normal Q-Q")


```

## 4.2 ANOVA Model Comparison

To test whether the addition of Demographic parameters and Lifestyle parameters have an effect on squared residuals of the model, an ANOVA F-test is utilized. The Demographic-only model was compared to the full model to assess the effect of Modifiable parameters; the THe Modifiable-only model was compared to the full model to assess the effect  


```{r}
# Output ANOVA Table 1
anova(m1_base, m3_full) |>
  kable(caption = "ANOVA Comparison: Demographics vs. Full Model", digits = 2) |>
  kable_styling(latex_options = "hold_position")

# Output ANOVA Table 2
anova(m2_life, m3_full) |>
  kable(caption = "ANOVA Comparison: Lifestyle vs. Full Model", digits = 2) |>
  kable_styling(latex_options = "hold_position")
```


**Addition of Demographic/Socioeconomic Risk Factors**
***H(null)**: \beta_{Age} = \beta_{Sex} =  \beta_{Region} = \beta_{Income} = \beta_{MaritalStatus} = \beta_{Urban/Rural} = \beta_{Education} = 0
***H(alt)**: At least one of  \beta_{Age}, \beta_{Sex}, \beta_{Region}, \beta_{Income}, \beta_{MaritalStatus}, \beta_{Urban/Rural},  \beta_{Education} ≠ 0
The The test statistic of 461.19 has a F distribution (df1 = 8, df2 = dummy) under the null hypothesis. The corresponding p value (apprxoimately 0) is clearly less than alpha value 0.05; this indicates that we can reject the null hypothesis and conclude that the demographic/socioeconomic parameters are not all equal to zero. Therefore, variation at least one of the socioeconomic/demographic factors seems to help explain the variation in the log(monthly premium); the addition of DRF helps reduce the squared residuals.

**Addition of Modifiable lifestyle choices**
***H(null)**: \beta_{Smoker} = \beta_{Alcohol} =  \beta_{bmi} = 0
**H(alt)**: At least one of \beta_{Smoker},  \beta_{Alcohol},  \beta_{bmi} ;  ≠ 0
The test statistic of 139.44 has a F distribution (df1 = 8, df2 = dummy)  under the null hypothesis. The corresponding p value (apprxoimately 0) is clearly less than alpha value 0.05; this indicates that we can reject the null hypothesis and conclude that the modifiable life choice parameters are not all equal to zero. Therefore, variation at least one of the modifiable life choices seems to help explain the variation in the log(monthly premium); the addition of MLC helps reduce the squared residuals.

**Overall Conclusion:**
The ANOVA tests conclude that the addition of MLC parameters to a DRF model may help explain the variation in the log(monthly premium). The same result was found when DRF parameters were added to a MLC model. However, as the degrees of freedom are not the same for each ANOVA test, the F value cannot be used directly to compare the impact of the added DRF/MLC variables.

## 4.3 Comparison of R^2 Values

```{r}
R21 <- summary(m1_base)$r.squared
R22 <- summary(m2_life)$r.squared
R23 <- summary(m3_full)$r.squared

dfr2 <- data.frame(
  Category = c("R\u00B2 Demographic", "R\u00B2 Lifestyle", " R\u00B2 Full Model"),
  R2_Value = c(R21, R22, R23)
)

kable(dfr2, caption = "R\u00B2 Values of Demographic, Lifestyle, and Full Model")

R231 <- R23 - R21
R232 <- R23 - R22

dfr2c <- data.frame(
  Category = c("ΔR\u00B2 (Full Model - Demographic)", "ΔR\u00B2 (Full Model - Lifestyle)"),
  R2_Value = c(R231, R232)
)

kable(dfr2c, caption = "ΔR\u00B2 of Demographic and Lifestyle Model Compared to Full Model")

```

**Change in Squared Residuals**
To compare the effect of adding lifestyle variables versus demographic variables, the R² values can instead be observed. R² values are extremely low for all three models, at 0.0249, 0.0342, and 0.0596 for the demographic, lifestyle and full model respectively (Table X). This indicates that all three models do not have a strong fit for the data. 

The effect of adding lifestyle variables versus demographic variables can be assed by comparing the change in R² value of the lifestyle and demographic model with the full model. The demographic-only model has a R² that is 0.0347 less than the full model (addition of lifestyle variables), while the lifestyle model has an R² that is 0.0255 less than the full model (addition of demographic variables). This indicates that the addition of the lifestyle variables has a more substantial effect on reducing the squared residuals of the model. 

## 4.4 Regression Coefficients
Table 3 presents the coefficients for the Full Model. Since the response is log-transformed, coefficients ($\beta$) can be interpreted as an approximate $100 \times \beta \%$ change in the premium for small values.

```{r}
# Tidy results for the main lifestyle factors
tidy(m3_full) |>
  filter(str_detect(term, "bmi|smoker|alcohol|age:smoker")) |>
  mutate(
    term = str_replace_all(term, "smokerCurrent", "Smoker: Current"),
    term = str_replace_all(term, "smokerFormer", "Smoker: Former"),
    term = str_replace_all(term, "alcohol_freqWeekly", "Alcohol: Weekly"),
    term = str_replace_all(term, "alcohol_freqOccasional", "Alcohol: Occasional"),
    term = str_replace_all(term, "alcohol_freqDaily", "Alcohol: Daily"),
    term = str_replace_all(term, "age:smokerCurrent", "Interaction: Age * Smoker"),
    term = str_replace_all(term, "bmi", "BMI"),
    estimate = round(estimate, 5),
    p.value = scales::pvalue(p.value)
  ) |>
  select(Predictor = term, Estimate = estimate, `Std. Error` = std.error, `P-Value` = p.value) |>
  kable(caption = "Coefficients for Modifiable Lifestyle Factors (Model 2)") |>
  kable_styling(latex_options = "hold_position")
```

Key Findings: 

1. Smoking: Being a "Current Smoker" has a large positive coefficient ($0.221$), indicating a substantial premium increase compared to never-smokers, holding other factors constant.

2. BMI: The coefficient for BMI is positive and statistically significant ($p < 0.001$). Higher body mass index is reliably associated with higher insurance costs.

3. Alcohol: The p-values for alcohol frequency are $> 0.05$. This suggests that, unlike smoking and BMI, self-reported alcohol frequency is not a statistically significant predictor of insurance premiums in this dataset.

4. Interaction (Age $\times$ Smoker): The significant interaction term ($p = 0.002$) confirms our EDA finding. The slope of the cost curve is steeper for smokers. This means the dollar-cost of smoking increases as the individual ages.


# 5. Discussion

**5.1 Locus of Control: Lifestyle Matters More** 
Our analysis reveals a crucial insight for consumers: Modifiable Lifestyle Choices (MLC) are stronger predictors of insurance premiums than Baseline Demographic Factors (DRF). Holding other variables constant, lifestyle choices explained ~3.4% of the variance. In contrast, demographics explained ~2.5% of the variance. This suggests that the factors driving costs up (Smoking, BMI) are within the individual's control.

**5.2 The "Double Penalty" of Smoking** 
The identification of a statistically significant interaction between Age and Smoking provides a strong argument against smoking amongst old individuals. An older smoker does not face the same monetary consequences as a younger smoker in terms of monthly premium; they pay a more significant fee.

**5.3 Critique of Methodology and Data Validity** 
Perhaps the most significant finding is what the model failed to explain. Our Full Model achieved an R² of only ~0.06. This implies 94% of the variance in insurance premiums is unexplained. This is likely due to Omitted Variable Bias [2]; premiums are heavily influenced by realized medical risk (e.g., Cancer history, Chronic conditions) which were absent from this dataset.

Additionally, the Data Reliability must be critiqued. For instance, the lack of significance for Alcohol Frequency may stem from reporting bias, as individuals often under-report substance use in surveys.

**5.4 Alternative Techniques & Future Work** 
While Linear Regression provided clear interpretability, the heavy tails in our residuals suggest that Quantile Regression or Generalized Linear Models (GLM) with a Gamma distribution might handle the high-cost outliers more effectively. Future research should prioritize obtaining real-world claims data containing specific medical history variables to improve predictive power.

# 6. References

[1] Kaggle. (2024). Medical Insurance Cost Prediction Dataset.
[2] Wooldridge, J. M. (2015). Introductory Econometrics.
[3] UVa Library. (2023). Hierarchical Linear Regression. https://library.virginia.edu/data/articles/hierarchical-linear-regression
[4] American Medical Association. (2024). National Health Expenditure Data. https://www.ama-assn.org/
[5] Bureau of Labor Statistics. (2023). Consumer Expenditure Survey: Health Insurance.
[6] HealthCare.gov. (2024). How Insurance Companies Set Health Premiums. https://www.healthcare.gov/how-plans-set-your-premiums/
[7] Kaiser Family Foundation. (2023). Employer Health Benefits Survey. https://www.kff.org/