---
title: "Determinants of Insurance Premiums: Lifestyle Choices vs. Demographic Factors"
author: "Varenya Katta, Raphael Karamagi, Tim Jun, Oat Twinvitoo"
format: pdf
geometry: margin=1in
header-includes:
  - \usepackage{float}
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(tidyverse)
library(knitr)
library(kableExtra)
library(gridExtra)
library(broom)

# Load data
insurance_raw <- read.csv("data/medical_insurance.csv")

# 1. Define all relevant columns
relevant_cols <- c(
  "annual_premium", "age", "sex", "education", "income", 
  "marital_status", "household_size", "region", "urban_rural", 
  "bmi", "smoker", "alcohol_freq"
)

# 2. Select Columns & Filter Zeros
insurance <- insurance_raw |>
  select(all_of(relevant_cols)) |>
  filter(annual_premium > 0)

# 3. Factor Levels 
insurance <- insurance |>
  mutate(
    smoker = factor(smoker, levels = c("Never", "Former", "Current")),
    alcohol_freq = factor(alcohol_freq, levels = c("None", "Occasional", "Weekly", "Daily")),
    sex = factor(sex),
    education = factor(education),
    marital_status = factor(marital_status),
    region = factor(region),
    urban_rural = factor(urban_rural)
  )

insurance <- na.omit(insurance)
```

# 1. Introduction
In 2023, healthcare spending in the United States increased by 7.5%, totaling 4.9 trillion USD [6]. This parallel increase in healthcare spending and health insurance costs has exceeded the growth rates of wages and inflation, making it increasingly difficult for individuals to afford coverage [5].

However, health insurance costs are not uniform; they vary significantly from one individual to another based on risk profiles. Legal frameworks allow insurers to adjust premiums based on specific factors such as age, location, and tobacco use [4]. Yet, individuals with pre-existing health risks or specific lifestyle habits may self-select into more expensive plans. We can broadly categorize the factors driving healthcare costs into two distinct groups: Modifiable Lifestyle Choices (MLC) (behaviors an individual can change) and Baseline Demographic/Socioeconomic Factors (DRF) (fixed characteristics).

While many studies investigate these variables individually, this study attempts to fill a research gap by comparing the relative explanatory power of these two categories using a large-scale dataset. Specifically, does one category—lifestyle or demographics—have a more substantial effect on insurance cost variance? The results have implications for both policy and personal finance: if costs are driven by MLC, individuals can lower premiums by modifying behavior; if driven by DRF, systemic policy changes may be required.

Research Question: **To what extent can variation in insurance premiums be attributed to Baseline Demographic Factors (DRF) versus Modifiable Lifestyle Factors (MLC), and does the financial burden of smoking compound with age?**

Data Description We utilize the Medical Insurance Cost Prediction dataset from Kaggle [1], comprising approximately 100,000 observations. The constituent data was derived from CDC, HHS, and CMS logic [1]. The response variable is the Annual Premium (USD). We analyze two sets of predictors:

Modifiable Lifestyle Choices (MLC): Smoking status (Current, Former, Never), Alcohol consumption frequency, and Body Mass Index (BMI).

Baseline Demographics (DRF): Age, Sex, Education level, Income, Marital Status, and Household Size.



# 2. Exploratory Data Analysis

## 2.1 Summary Statistics

Table 1 summarizes the central tendencies of the continuous variables. The substantial difference between the median (463.58 USD) and mean (582.32 USD) premium indicates a right-skewed distribution, suggesting the presence of high-cost outliers.
```{r}
#| label: tbl-summary
insurance |>
  summarise(
    `Average Age` = mean(age),
    `Average BMI` = mean(bmi),
    `Median Premium` = median(annual_premium),
    `Mean Premium` = mean(annual_premium),
    `Max Premium` = max(annual_premium)
  ) |>
  kable(caption = "Summary of Numerical Variables", digits = 2) |>
  kable_styling(latex_options = "hold_position")
```

## 2.2 Distribution and Transformation

Linear regression assumes approximately normal residuals with constant variance. When the response variable is heavily right-skewed, these assumptions are typically violated. As shown in Figure 1 (Left), the raw annual premium distribution exhibits substantial right-skewness. We therefore applied a log-transformation [2], ln(Premium), which produces a more symmetric distribution (Figure 1, Right). This transformation typically improves residual normality and homoscedasticity (equal variances), which we verify in Section 4.1.

```{r}
# Fit Models
m1_base <- lm(log(annual_premium) ~ age + sex + education + income + marital_status + household_size + region + urban_rural, data = insurance)

m2_life <- lm(log(annual_premium) ~ smoker + alcohol_freq + bmi, data = insurance)

m3_full <- lm(log(annual_premium) ~ age + sex + education + income + marital_status + household_size + region + urban_rural + bmi + smoker + alcohol_freq + age*smoker, data = insurance)
```

```{r}
#| label: fig-distribution
#| fig-cap: "Distribution of annual premiums before and after log-transformation"
#| #| fig-pos: 'H'

p1 <- ggplot(insurance, aes(x = annual_premium)) +
  geom_histogram(bins = 50, fill = "orange", color = "white") +
  labs(title = "Raw Distribution (Skewed)", 
       x = "Annual Premium ($)", y = "Count") +
  theme_minimal()

p2 <- ggplot(insurance, aes(x = log(annual_premium))) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(title = "Log-Transformed (Normal)", 
       x = "Log(Premium)", y = "Count") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

## 2.3 Visualizing the "Smoking Penalty" and Interaction
We hypothesized that the cost burden of smoking might vary by age. Figure 2 plots Age against Log-Premium, stratified by smoking status.

The Green Line (Current Smokers) shows a higher intercept, indicating a higher baseline cost. More importantly, the slope appears steeper than that of the Red Line (Never Smokers). This suggests a potential interaction effect: the financial penalty for smoking compounds as the individual ages. We will test this formally in our regression models.

```{r}
#| label: fig-interaction
#| fig-pos: 'H'
#| fig-cap: "Log-premium by age, stratified by smoking status. The steeper slope for current smokers suggests an interaction effect."


ggplot(insurance, aes(x = age, y = log(annual_premium), color = smoker)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", se = FALSE, size = 1.5) +
  labs(title = "Interaction: Smoking Costs Increase Faster with Age",
       x = "Age of Individual", 
       y = "Log(Annual Premium)", 
       color = "Smoking Status") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")
```


# 3. Methodology

To isolate the variance explained by lifestyle choices versus demographics, we employ Hierarchical Multiple Linear Regression [3]. This approach allows us to enter variables in blocks to determine the incremental variance explained by each category. We utilize Analysis of Variance (ANOVA) using Partial F-tests to determine if the addition of a block of variables significantly reduces the Sum of Squared Errors (SSE) compared to a reduced model.

We fit three distinct models:

Model 1: Baseline Demographics Accounts only for fixed factors the individual cannot easily change. $$ \ln(Y) = \beta_0 + \beta_{Age} + \beta_{Sex} + \beta_{Region} + \beta_{Income} + \dots + \epsilon $$

Model 2: Lifestyle Only Accounts only for modifiable behaviors. $$ \ln(Y) = \beta_0 + \beta_{Smoker} + \beta_{Alcohol} + \beta_{BMI} + \epsilon $$

Model 3: Full Model (Demographics + Lifestyle) Combines both categories and includes the interaction term observed in EDA. $$ \ln(Y) = \text{Model 1} + \text{Model 2} + (\beta_{Age} \times \beta_{Smoker}) + \epsilon $$

Justification and Assumptions

We chose Ordinary Least Squares (OLS) regression on the log-transformed outcome. OLS provides highly interpretable coefficients, allowing us to quantify the precise percentage change in premiums associated with each lifestyle factor. The log-transformation addresses the right-skewness identified in EDA.

To ensure our results are valid, we will assess four key assumptions in the Results section:

1. Linearity: The relationship between predictors and the log-response is linear
2. Independence: We assume observations are independent (i.e., one individual's premium doesn't influence another's)
3. Equal Variance: The variance of residuals is constant across fitted values
4. Normality: The residuals are normally distributed

# 4. Results

## 4.1 Diagnostics and Assumptions Checking
To ensure the validity of our inference, we examined the residual plots for the full Model (Figure 3).

- Linearity & Equal Variance: The Residuals vs Fitted plot (Left) shows a random cloud of points around zero with no distinct pattern (e.g., no funnel shape). This indicates that the assumptions of linearity and constant variance are reasonably satisfied.

- Normality of Residuals: (Right) shows the Normal Q-Q plot. The points (circles) adhere closely to the diagonal line for a large number of observations. However, the tails deviate upward. This  behavior indicates that our model slightly underestimates the premiums for the most extreme, high-cost outliers. Given the large sample size ($N \approx 100,000$), the Central Limit Theorem ensures that our coefficient estimates remain unbiased despite this deviation [2].

- Independence: One key source of this dataset is the National Health Interview survey; this survey is conducted on a voluntary basis. Despite a small risk of self selection bias, due to the large sample size, it's reasonable to assume independence for our model.


```{r}
#| label: fig-diagnostics
#| #| fig-pos: 'H'
#| #| fig-cap: "Diagnostic plots for the Full Model. Left: Residuals vs. Fitted values. Right: Normal Q-Q plot."


par(mfrow = c(1, 2))
plot(m3_full, which = 1, main = "Residuals vs Fitted")
plot(m3_full, which = 2, main = "Normal Q-Q")


```

## 4.2 ANOVA Model Comparison

To test whether the addition of specific variable categories significantly improves the model, we utilized ANOVA F-tests.


```{r}
# Output ANOVA Table 1
anova(m1_base, m3_full) |>
  kable(caption = "ANOVA Comparison: Demographics vs. Full Model", digits = 2) |>
  kable_styling(latex_options = "hold_position")

# Output ANOVA Table 2
anova(m2_life, m3_full) |>
  kable(caption = "ANOVA Comparison: Lifestyle vs. Full Model", digits = 2) |>
  kable_styling(latex_options = "hold_position")
```

**Testing the Addition of Demographic Factors**

$H_0: \beta_{Age} = \beta_{Sex} = \beta_{Region} = \beta_{Income} = \dots = 0$
$H_A: \text{At least one demographic coefficient } \neq 0$

We compared the Lifestyle-only model against the Full model to see if adding Demographics significantly reduced the squared residuals. The F-test yielded an F-statistic of 461.19 and a p-value $< 0.001$. We reject the null hypothesis. Therefore, demographic factors provide statistically significant explanatory power.

**Testing the Addition of Modifiable Lifestyle Factors** 

$H_0: \beta_{Smoker} = \beta_{Alcohol} = \beta_{BMI} = 0$
$H_A: \text{At least one lifestyle coefficient } \neq 0$

We compared the Demographic-only model against the Full model to see if adding Lifestyle factors improved the fit. The F-test yielded an F-statistic of 128.88 and a p-value $< 0.001$. We reject the null hypothesis. This confirms that modifiable choices (Smoking, BMI) also significantly explain variation in premiums.

## 4.3 Comparison of R² Values

While ANOVA tests for statistical significance, comparing R² values allows us to assess the practical explanatory power (effect size).

```{r}
R21 <- summary(m1_base)$r.squared
R22 <- summary(m2_life)$r.squared
R23 <- summary(m3_full)$r.squared

dfr2 <- data.frame(
  Category = c("R\u00B2 Demographic", "R\u00B2 Lifestyle", " R\u00B2 Full Model"),
  R2_Value = c(R21, R22, R23)
)

kable(dfr2, caption = "R\u00B2 Values of Demographic, Lifestyle, and Full Model")

R231 <- R23 - R21
R232 <- R23 - R22

dfr2c <- data.frame(
  Category = c("ΔR\u00B2 (Full Model - Demographic)", "ΔR\u00B2 (Full Model - Lifestyle)"),
  R2_Value = c(R231, R232)
)

kable(dfr2c, caption = "ΔR\u00B2 of Demographic and Lifestyle Model Compared to Full Model")

```

The R² values are generally low across all models (0.0249 to 0.0596), indicating that the variables available in this dataset explain only a small fraction of the total variance in premiums.

However, we can compare the relative impact of the two categories. Adding Lifestyle variables to the demographic model increased the R² by roughly 0.0347. Conversely, adding Demographic variables to the lifestyle model increased the R² by only 0.0255. This suggests that modifiable lifestyle factors have a stronger marginal effect on predicting insurance costs than fixed demographic factors.


## 4.4 Regression Coefficients
Table 5 presents the coefficients for the Full Model. Since the response is log-transformed, coefficients ($\beta$) can be interpreted as an approximate $100 \times \beta \%$ change in the premium for small values.

```{r}
# Tidy results for the main lifestyle factors
tidy(m3_full) |>
  filter(str_detect(term, "bmi|smoker|alcohol|age:smoker")) |>
  mutate(
    term = str_replace_all(term, "smokerCurrent", "Smoker: Current"),
    term = str_replace_all(term, "smokerFormer", "Smoker: Former"),
    term = str_replace_all(term, "alcohol_freqWeekly", "Alcohol: Weekly"),
    term = str_replace_all(term, "alcohol_freqOccasional", "Alcohol: Occasional"),
    term = str_replace_all(term, "alcohol_freqDaily", "Alcohol: Daily"),
    term = str_replace_all(term, "age:smokerCurrent", "Interaction: Age * Smoker"),
    term = str_replace_all(term, "bmi", "BMI"),
    estimate = round(estimate, 5),
    p.value = scales::pvalue(p.value)
  ) |>
  select(Predictor = term, Estimate = estimate, `Std. Error` = std.error, `P-Value` = p.value) |>
  kable(caption = "Coefficients for Modifiable Lifestyle Factors (Model 2)") |>
  kable_styling(latex_options = "hold_position")
```

Key Findings: 

1. Smoking: Being a "Current Smoker" has a large positive coefficient ($0.221$). Holding other factors constant, current smokers pay approximately 22% more in premiums than never-smokers at the baseline age.

2. BMI: The coefficient for BMI is positive and statistically significant ($p < 0.001$). Higher body mass index is reliably associated with higher insurance costs.

3. Alcohol: The p-values for alcohol frequency are $> 0.05$. This suggests that, unlike smoking and BMI, self-reported alcohol frequency is not a statistically significant predictor of insurance premiums in this dataset.

4. Interaction (Age $\times$ Smoker): The significant interaction term ($p = 0.002$) confirms our EDA finding. This positive coefficient ($\beta = 0.00087$)) means the premium gap between smokers and non-smokers widens as age increases. For every additional year of age, the "smoker penalty" increases.

# 5. Discussion

### 5.1 Locus of Control: Lifestyle Matters More
Our analysis reveals a crucial insight for consumers: Modifiable Lifestyle Choices (MLC) are stronger predictors of insurance premiums than Baseline Demographic Factors (DRF). The change in $R^2$ analysis demonstrated that lifestyle choices explained more unique variance ($\Delta R^2 \approx 0.035$) than demographics ($\Delta R^2 \approx 0.025$). This is empowering, as it suggests that the primary factors driving costs up (Smoking, BMI) are within the individual's control, rather than being determined solely by fixed characteristics like age or region.

### 5.2 The "Double Penalty" of Smoking
The identification of a statistically significant interaction between Age and Smoking provides a compelling financial incentive for cessation. A 50-year-old smoker does not just pay the standard "smoker fee"; they pay a compounded rate compared to a 20-year-old smoker. This suggests that the "return on investment" for quitting smoking increases as one gets older. Policy-makers could leverage this finding to target smoking cessation programs specifically toward middle-aged demographics, emphasizing the compounding financial savings.

###5.3 Critique of Methodology and Data Validity 
Perhaps the most significant finding is what the model failed to explain. Our Full Model achieved an $R^2$ of only ~0.06. This implies 94% of the variance in insurance premiums is unexplained. This is likely due to Omitted Variable Bias [2]; premiums are primarily influenced by realized medical risk (e.g., Cancer history, Chronic conditions, Prescription usage) which were absent from this dataset. While our specific variables (Smoking, Age) were statistically significant, they are not sufficient for accurate price prediction on their own.

### 5.4 Alternative Techniques & Future Work
While Linear Regression provided clear interpretability, the heavy tails in our residuals suggest that Quantile Regression or Generalized Linear Models (GLM) with a Gamma distribution might handle the high-cost outliers more effectively. Future research should prioritize obtaining real-world claims data containing specific medical history variables to drastically improve predictive power.


# 6. References

[1] Kaggle. (2024). Medical Insurance Cost Prediction Dataset.
[2] Wooldridge, J. M. (2015). Introductory Econometrics.
[3] UVa Library. (2023). Hierarchical Linear Regression. https://library.virginia.edu/data/articles/hierarchical-linear-regression
[4] HealthCare.gov. (2024). How Insurance Companies Set Health Premiums. https://www.healthcare.gov/how-plans-set-your-premiums/
[5] Kaiser Family Foundation. (2023). Employer Health Benefits Survey. https://www.kff.org/
[6] Centers for Medicare & Medicaid Services, Office of the Actuary. “National Health Expenditure (NHE) Fact Sheet: Historical NHE, 2023.” U.S. Department of Health & Human Services