---
title: "Determinants of Insurance Premiums: Lifestyle Choices vs. Demographic Factors"
author: "Varenya Katta, Raphael Karamagi, Tim Jun, Oat Twinvitoo"
format: pdf
geometry: margin=1in
header-includes:
  - \usepackage{float}
execute:
  echo: false
  warning: false
  message: false
  fig.width: 4
  fig.height: 3
---

```{r setup}
#| include: false
library(tidyverse)
library(knitr)
library(kableExtra)
library(gridExtra)
library(broom)

# Load data
insurance_raw <- read.csv("data/medical_insurance.csv")

# 1. Define all relevant columns
relevant_cols <- c(
  "annual_premium", "age", "sex", "education", "income", 
  "marital_status", "household_size", "region", "urban_rural", 
  "bmi", "smoker", "alcohol_freq"
)

# 2. Select Columns & Filter Zeros
insurance <- insurance_raw |>
  select(all_of(relevant_cols)) |>
  filter(annual_premium > 0)

# 3. Factor Levels 
insurance <- insurance |>
  mutate(
    smoker = factor(smoker, levels = c("Never", "Former", "Current")),
    alcohol_freq = factor(alcohol_freq, levels = c("None", "Occasional", "Weekly", "Daily")),
    sex = factor(sex),
    education = factor(education),
    marital_status = factor(marital_status),
    region = factor(region),
    urban_rural = factor(urban_rural)
  )

insurance <- na.omit(insurance)
```

# 1. Introduction

# 1.1 Context and Research Question

In 2023, healthcare spending in the United States increased by 7.5%, totaling 4.9 trillion USD [8]. This increase in healthcare spending paralells increases in health insurance costs. As of recent health insurance costs have exceeded the growth rates of wages and inflation, making it increasingly difficult for individuals to afford coverage [7].

However, health insurance vary from individual to individual; according to the Affordable Care Act, factors such as age, location and tobacco use are legally allowed to increase the cost of one's premium [6]. Further, individuals with other pre-existing health risks may also select costlier healthcare plans. Overall, we can categorize factors that potentially drive healthcare cost into two categories: Modifiable Lifestyle Choices (MLC) (behaviors an individual can change) and Baseline Demographic/Socioeconomic Factors (DRF) (fixed characteristics).

While previous studies have examined these variables individually, this study will attempt to bridge a gap by comparing the extent of variability attributable to each of these two categories: Does one category have a more substantial effect on insurance cost variance? The results may indicate how institutions and individuals can make healthcare more affordable. If insurance cost variance is heavily driven by MLC, customers can lower prices by modifying behaviors. However, if insurance cost is heavily driven by DRF, policy changes may be required to lower costs.

Research Question: **To what extent can variation in insurance premiums be attributed to Baseline Demographic Factors (DRF) versus Modifiable Lifestyle Factors (MLC), and does the financial burden of smoking compound with age?**

**Dataset:** We utilize the Medical Insurance Cost Prediction dataset from Kaggle [1]. The dataset contains approximately 100,000 observations including demographic details (Age, Sex, Income) and lifestyle indicators (Smoking status, BMI, Alcohol frequency). Constituent data was derived from three main sources: Centers for Disease Control and Prevention (CDC), U.S. Department of Health & Human Services (HHS), and Centers for Medicare & Medicaid Services (CMS). The response variable is the annual healthcare premium cost.

Modifiable Lifestyle Choices (MLC): Smoking status (Current, Former, Never), Alcohol consumption frequency, and Body Mass Index (BMI).

Baseline Demographics (DRF): Age, Sex, Education level, Income, Marital Status, and Household Size.

# 1.2 Exploratory Data Analysis

Table 1 provides summary statistics of the response variable (Annual Premium), namely, median, mean, minimum and maximum. As the mean premium, at 582.32 USD, is substantially higher than the median, at 463.58 USD, we can infer a right-skewed distribution. Further, the extremely high value of the maximum premium, at 10962.55 USD, especially in comparison to the minimum premium value of 211.67 USD, may imply the presence of high outliers. 

```{r}
insurance |>
  summarise(
    `Median Premium` = median(annual_premium),
    `Mean Premium` = mean(annual_premium),
    `Min Premium` = min(annual_premium),
    `Max Premium` = max(annual_premium)
  ) |>
  kable(caption = "Summary Statistics of Annual Premium Costs", digits = 2) |>
  kable_styling(latex_options = "hold_position")
```



# 2. Methodology

```{r}
# Fit Models
p3_full <- lm((annual_premium) ~ age + sex + education + income + marital_status + household_size + region + urban_rural + bmi + smoker + alcohol_freq, data = insurance)
```

```{r}
# Fit Models
m1_base <- lm(log(annual_premium) ~ age + sex + education + income + marital_status + household_size + region + urban_rural, data = insurance)

m2_life <- lm(log(annual_premium) ~ smoker + alcohol_freq + bmi, data = insurance)

m3_full <- lm(log(annual_premium) ~ age + sex + education + income + marital_status + household_size + region + urban_rural + bmi + smoker + alcohol_freq + age*smoker, data = insurance)
```

To isolate the variance explained by lifestyle choices versus demographics, we employ Hierarchical Multiple Linear Regression [3]. This approach allows us to enter variables in blocks to determine the unique variance explained by each category. We utilize Analysis of Variance (ANOVA) using Partial F-tests to determine if the addition of a block of variables significantly reduces the Sum of Squared Errors (SSE) compared to a reduced model.

We fit three distinct models:

Model 1: Baseline Demographics Accounts only for fixed factors the individual cannot easily change. $$ \ln(Y) = \beta_0 + \beta_{Age} + \beta_{Sex} + \beta_{Region} + \beta_{Income} + \dots + \epsilon $$

Model 2: Lifestyle Only Accounts only for modifiable behaviors. $$ \ln(Y) = \beta_0 + \beta_{Smoker} + \beta_{Alcohol} + \beta_{BMI} + \epsilon $$

Model 3: Full Model (Demographics + Lifestyle) Combines both categories and includes the interaction term observed in EDA. $$ \ln(Y) = \text{Model 1} + \text{Model 2} + \epsilon $$

While ANOVA tests for statistical significance, it does not quantify effect size. Therefore, we also compare the change in Adjusted R² between the partial models (DRF vs. Full and MLC vs. Full). This allows us to assess which set of variables has a more substantial practical effect on reducing the squared residuals.

# 2.1 Log Transformation
The use of a log transformation is to address the requirement of normality in the residuals required by Least Squares Regression [2]. When a full model is fit without it, its associated residual histogram is right skewed (Figure 1). To address this, we applied a  log-transformation [2] to the outcome variable: annual premium. Doing so produced more symmetric residual distribution (Figure 2), and allowed for the normality requirement to be satisfied. Other models (DRF and MLC) also showed roughly normal distributions when log was applied to the outcome variable (Figure 2). 

```{r}
#| fig-pos: 'H'
#| label: fig-rawresiduals


p3 <- ggplot(data = p3_full, mapping = aes(x = p3_full$residuals)) + 
  geom_histogram(bins = 50, fill = "orange", color = "white") +
  labs(title = "Raw Distribution of Residuals (Skewed)", 
       x = "Residual Annual Premium ($)", y = "Count") +
  theme_minimal()

grid.arrange(p3, ncol = 1)
```

```{r}
#| fig-pos: 'H'
#| label: fig-residuals


m1 <- ggplot(data = m1_base, mapping = aes(x = m1_base$residuals)) + 
  geom_histogram(bins = 50, fill = "orange", color = "white") +
  labs(title = "Distribution of Residuals for DRF Model", 
       x = "Residual Log(Annual Premium) ($)", y = "Count") +
  theme_minimal()
m2 <- ggplot(data = m2_life, mapping = aes(x = m2_life$residuals)) + 
  geom_histogram(bins = 50, fill = "orange", color = "white") +
  labs(title = "Distribution of Residuals for MLC Model", 
       x = "Residual Log(Annual Premium) ($)", y = "Count") +
  theme_minimal()
m3 <- ggplot(data = m3_full, mapping = aes(x = m3_full$residuals)) + 
  geom_histogram(bins = 50, fill = "orange", color = "white") +
  labs(title = " Distribution of Residuals for Full Model", 
       x = "Residual Log(Annual Premium) ($)", y = "Count") +
  theme_minimal()


grid.arrange(m1, m2, m3, ncol = 1)
```

# 2.2 Justification and Other Assumptions

We chose Ordinary Least Squares (OLS) regression on the log-transformed outcome. OLS provides highly interpretable coefficients, allowing for the quantification the changes in premiums associated with each lifestyle and demographic variable.

To ensure our results are valid, we will assess four key assumptions in the Results section:

1. Linearity: The relationship between predictors and the log-response is linear
2. Independence: We assume observations are independent (i.e., one individual's premium doesn't influence another's)
3. Equal Variance: The variance of residuals is constant across fitted values
4. Normality: The residuals are normally distributed

# 3. Results

## 3.1 Diagnostics and Assumptions Checking
To ensure the validity of our inference, we examined the residual plots for the full Model (Figure 3).

- Linearity & Equal Variance: The Residuals vs Fitted plot (Left) shows a random cloud of points around zero with no distinct pattern (e.g., no funnel shape). This indicates that the assumptions of linearity and constant variance are reasonably satisfied.

- Independence: One key source of this dataset [1] is the National Health Interview survey; this survey is conducted on a voluntary basis. Consequently, there is a self-selection bias that may affect the ability to fully assume independence for this study.

- Normality of Residuals: As shown earlier in the residual histograms, the normality of the residuals for each of the three models is roughly fulfilled. Further, figure 4 (Right) shows the Normal Q-Q plot for the full model. The points (circles) adhere closely to the diagonal line for the majority of observations. However, the tails deviate upward. This "heavy-tailed" behavior indicates that our model slightly underestimates the premiums for the most extreme, high-cost outliers. Given the large sample size ($N \approx 100,000$), the Central Limit Theorem ensures that our coefficient estimates remain unbiased despite this deviation [2].


```{r}
#| label: fig-diagnostics
#| fig-pos: 'H'
#| fig-cap: "Diagnostic plots for the Full Model. Left: Residuals vs. Fitted values. Right: Normal Q-Q plot."


par(mfrow = c(1, 2))
plot(m3_full, which = 1, main = "Residuals vs Fitted")
plot(m3_full, which = 2, main = "Normal Q-Q")


```

## 3.2 ANOVA Model Comparison

To test whether the addition of specific variable categories significantly improves the model, we utilized ANOVA F-tests.


```{r}
# Output ANOVA Table 1
anova(m1_base, m3_full) |>
  kable(caption = "ANOVA Comparison: Demographics vs. Full Model", digits = 2) |>
  kable_styling(latex_options = "hold_position")

# Output ANOVA Table 2
anova(m2_life, m3_full) |>
  kable(caption = "ANOVA Comparison: Lifestyle vs. Full Model", digits = 2) |>
  kable_styling(latex_options = "hold_position")
```

**Testing the Addition of Demographic Factors**

**H(null):** $\beta_{Age} = \beta_{Sex} =  \beta_{Region} = \beta_{Income} = \beta_{MaritalStatus} = \beta_{Urban/Rural} = \beta_{Education}  0$

**H(alt):** At least one of  $\beta_{Age}, \beta_{Sex}, \beta_{Region}, \beta_{Income}, \beta_{MaritalStatus}, \beta_{Urban/Rural},  \beta_{Education} ≠ 0$


We compared the Lifestyle-only model against the Full model to see if adding Demographics significantly reduced the squared residuals. The F-test yielded an F-statistic of 461.19 and a p-value <0.001. We reject the null hypothesis. This confirms that demographic factors (Age, Income, etc.) do provide statistically significant explanatory power.


**Testing the Addition of Modifiable Lifestyle Factors** 

**H(null)**: $\beta_{Smoker} = \beta_{Alcohol} =  \beta_{BMI} = 0$

**H(alt)**: At least one of $\beta_{Smoker},  \beta_{Alcohol},  \beta_{BMI}  ≠ 0$

We compared the Demographic-only model against the Full model to see if adding Lifestyle factors improved the fit. The F-test yielded an F-statistic of 128.88 and a p-value <0.001. We reject the null hypothesis. This confirms that modifiable choices (Smoking, BMI) also significantly explain variation in premiums.

The ANOVA tests conclude that the addition of MLC parameters to a DRF model helps explain the variation in the log(annual premium). However, as the degrees of freedom are not the same for each ANOVA test, the F value cannot be used directly to compare the practical impact of the added variables.

## 3.3 Comparison of R² Values

While ANOVA tests for statistical significance, comparing R² values allows us to assess the practical explanatory power (effect size).

```{r}
R21 <- summary(m1_base)$r.squared
R22 <- summary(m2_life)$r.squared
R23 <- summary(m3_full)$r.squared

dfr2 <- data.frame(
  Category = c("R\u00B2 Demographic", "R\u00B2 Lifestyle", " R\u00B2 Full Model"),
  R2_Value = c(R21, R22, R23)
)

kable(dfr2, caption = "R\u00B2 Values of Demographic, Lifestyle, and Full Model")

R231 <- R23 - R21
R232 <- R23 - R22

dfr2c <- data.frame(
  Category = c("ΔR\u00B2 (Full Model - Demographic)", "ΔR\u00B2 (Full Model - Lifestyle)"),
  R2_Value = c(R231, R232)
)

kable(dfr2c, caption = "ΔR\u00B2 of Demographic and Lifestyle Model Compared to Full Model")

```

The R² values are generally low across all models (0.0249 to 0.0596), indicating that the variables available in this dataset explain only a small fraction of the total variance in premiums.

However, we can compare the relative impact of the two categories. Adding Lifestyle variables to the demographic model increased the R² by roughly 0.0347. Conversely, adding Demographic variables to the lifestyle model increased the R² by only 0.0255. This suggests that modifiable lifestyle factors have a stronger marginal effect on predicting insurance costs than fixed demographic factors.


## 3.4 Regression Coefficients
Table 5 presents the coefficients for the Full Model. Since the response is log-transformed, coefficients ($\beta$) can be interpreted as an approximate $100 \times \beta \%$ change in the premium for small values.

```{r}
# Tidy results for the main lifestyle factors
tidy(m3_full) |>
  filter(str_detect(term, "bmi|smoker|alcohol|age:smoker")) |>
  mutate(
    term = str_replace_all(term, "smokerCurrent", "Smoker: Current"),
    term = str_replace_all(term, "smokerFormer", "Smoker: Former"),
    term = str_replace_all(term, "alcohol_freqWeekly", "Alcohol: Weekly"),
    term = str_replace_all(term, "alcohol_freqOccasional", "Alcohol: Occasional"),
    term = str_replace_all(term, "alcohol_freqDaily", "Alcohol: Daily"),
    term = str_replace_all(term, "bmi", "BMI"), 
    estimate = round(estimate, 5),
    p.value = scales::pvalue(p.value)
  ) |>
  select(Predictor = term, Estimate = estimate, `Std. Error` = std.error, `P-Value` = p.value) |>
  kable(caption = "Coefficients for Full Model (Model 3)") |>
  kable_styling(latex_options = "hold_position")
```

Key Findings: 

1. Smoking: Being a "Current Smoker" has a large positive coefficient ($0.221$). Holding other factors constant, current smokers pay approximately 22% more in premiums than never-smokers at the baseline age.

2. BMI: The coefficient for BMI is positive and statistically significant ($p < 0.001$). Higher body mass index is reliably associated with higher insurance costs.

3. Alcohol: The p-values for alcohol frequency are $> 0.05$. This suggests that, unlike smoking and BMI, self-reported alcohol frequency is not a statistically significant predictor of insurance premiums in this dataset.

# 4. Discussion

## 4.1 Interpretation of Results
Initial analysis with hierarchical multiple linear regression reveals that lifestyle choices and demographic factors affect the log(annual insurance). However, further examination into R² values revealed that the most of the variation in log(annual insurance) could not be explained with models generated with MLC variables alone, DRF variables alone, and both MLC and DRF variables. Notably, the full model has a R² value of 0.06, indicating that 6% of the variation of the log(annual insurance) can be explained by the regression model. The DRF model has an R² value of 0.025, meaning 2.5% of the variation can be explained by the model and MLC alone has an R² value of 0.034, indicating that 3.4% of the variation can be explained by the model. 
Comparing the change in R² value between the full vs. DRF model and full vs. MLC model. We find that the difference between the full and DRF model ( ΔR² = 0.0347) is higher than that of the difference between the full and MLC model ( ΔR² = 0.0254). This indicates that the addition of lifestyle variables to a DRF model has a greater impact in minimizing R² than adding demographic variables to an MLC model.
The low R² value of this data indicates that a substantial amount of log(annual premium) variation is attributable to factors not present in the dataset. This may include variables such as deductible amounts, coverage levels, provider networks, and specific insurer pricing strategies.
When observing the slope coefficients of the MLC model, we found that smoking was associated with a high slope value. Current smokers have a log(annual premium) of 0.22 higher than those who have never smoked. BMI similarly showed a positive relationship with log(annual premium), with an increase in one BMI point being associated with an increase in log(annual premium) of 0.56. In contrast with these two variables, alcohol consumption did not seem to have a significant effect on log(annual premium) due to p values (for occasional, weekly, and daily drinking) all greater than 0.01. Our findings align with the policies laid out in the affordable care act: while smoking may be used to adjust insurance cost, BMI and alcohol consumption are not considered. 

## 4.2 Methodological Limitations & Reccomendations
Several issues limit the strength of these conclusions. First, some factors that are known to have been able to legally affect healthcare costs, such as plan types, are not included in the dataset. Therefore, these significant variables cannot be controlled for and cannot be assessed.  Second, some of the data sources such as the National Health Interview Survey collects self reported data. Relying upon such data for factors such as BMI, smoking and alcohol consumption may be prone to error as individuals may want to underreport such stigmatized behaviors. Third, the model underestimates premiums for the individuals paying the most for insurance, as shown in the Q-Q plot.  Fourth, as the variables selected do not represent the totality of Modifiable Lifestyle Choices and Demographics, a general conclusion about these two factors may be inappropriate. 
To improve the study, more variables representing MLC and DRF should be included. It may also be helpful to test the interaction effects between MLC and DRF variables. Further, more categories for alcohol consumption may be helpful. Currently, there is no distinction between an individual who has quit alcohol recently and another who has quit alcohol for a long duration of time. There may exist a distinction in insurance premium between these two groups. Finally, a dataset that does not rely on self-reported data should be found.

## 4.3 Implications
Our results reveal that lifestyle choices and demographics contribute to the cost of insurance. The addition of lifestyle choice variables seem to help explain variability in log(annual insurance cost) more significantly when demographic factors were added. Despite this, the examined variables seem to explain very little variation in log(annual insurance costs), as indicated by the low R² value. Simultaneously, this implies that other variables not investigated in this study may better explain variations in health insurance. Consequently, further study is needed into the explanatory effect of modifiable lifestyle choice variables vs. demographic variables on insurance cost.

# 5. References

[1] Kaggle. (2024). Medical Insurance Cost Prediction Dataset.
[2] Wooldridge, J. M. (2015). Introductory Econometrics.
[3] UVa Library. (2023). Hierarchical Linear Regression. https://library.virginia.edu/data/articles/hierarchical-linear-regression
[4] American Medical Association. (2024). National Health Expenditure Data. https://www.ama-assn.org/
[5] Bureau of Labor Statistics. (2023). Consumer Expenditure Survey: Health Insurance.
[6] HealthCare.gov. (2024). How Insurance Companies Set Health Premiums. https://www.healthcare.gov/how-plans-set-your-premiums/
[7] Kaiser Family Foundation. (2023). Employer Health Benefits Survey. https://www.kff.org/
[8] Centers for Medicare & Medicaid Services, Office of the Actuary. “National Health Expenditure (NHE) Fact Sheet: Historical NHE, 2023.” U.S. Department of Health & Human Services
